---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# rFIA: Unlocking the FIA Database in R

<!-- badges: start -->
<!-- badges: end -->

The goal of `rFIA` is to increase the accessibility and use of the USFS Forest Inventory and Analysis (FIA) Database by providing a user-friendly, open source platform to easily query and analyze FIA Data. Designed to accommodate a wide range of potential user objectives, `rFIA` simplifies the estimation of forest variables from the FIA Database and allows all R users (experts and newcomers alike) to unlock the flexibility and potential inherent to the Enhanced FIA design.

Specifically, `rFIA` improves accessibility to the spatio-temporal estimation capacity of the FIA Database by producing space-time indexed summaries of forest variables within user-defined population boundaries. Direct integration with other popular R packages (e.g., `dplyr`, `sp`, and `sf`) facilitates efficient space-time query and data summary, and supports common data representations and API design. The package implements design-based estimation procedures outlined by Bechtold & Patterson (2005), and has been validated against estimates and sampling errors produced by EVALIDator. Current development is focused on the implementation of spatially-enabled model-assisted estimators to improve population, change, and ratio estimates.

You can download subsets of the FIA Database at the FIA Datamart: https://apps.fs.usda.gov/fia/datamart/CSV/datamart_csv.html


## Installation

You can install the released version of `rFIA` from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("rFIA")
```

Alternatively, you can install the development version from GitHub:
```r
devtools::install_github('hunter-stanke/rFIA')
```

## Functionality

|`rFIA` Function  | Description                                                          |
|---------------- |----------------------------------------------------------------------|
|`animateFIA`     | Produce animated plots of spatial FIA summaries                      |
|`biomass`        | Estimate volume, biomass, & carbon stocks of standing trees          |
|`clipFIA`        | Spatial & temporal queries                                           |
|`diversity`      | Estimate species diversity                                           |
|`dwm`            | Estimate volume, biomass, and carbon stocks of down woody material   |
|`growMort`       | Estimate recruitment, mortality, and harvest rates                   |
|`invasive`       | Estimate areal coverage of invasive species                          |
|`readFIA`        | Load FIA database into R environment                                 |
|`standStruct`    | Estimate forest structural stage distributions                       |
|`tpa`            | Estimate abundance of standing trees (TPA & BAA)                     |
|`vitalRates`     | Estimate live tree growth rates                                      |


## Example Usage

#### _Load FIA Data into R_


The first step to using `rFIA` is to download subsets of the FIA Database from the FIA Datamart. Download a state subset by clicking the state of interest here: https://apps.fs.usda.gov/fia/datamart/CSV/datamart_csv.html. 

Note: given the massive size of the full FIA Database, users are cautioned to only download the subsets containing their region of interest.

Once downloaded, simply unzip the files in thier own directory (all files in the same directory, with no non-FIA files present), and load the database into R using `readFIA`:
```{r eval = FALSE}
## Load FIA Data from a local directory
db <- readFIA('/path/to/your/directory/')
```
By default, `readFIA` only loads the portions of the database required to 
produce summaries with `rFIA` functions (`common = TRUE`)

**But what if I want to load multiple states worth of FIA data into R?** No problem! Simply unzip the files from each state download and place them in the same directory. Point `readFIA` to this directory, and it will load and merge the states into a single `FIA.Database` object. This will allow you to use other `rFIA` functions to produce estimates within polygons which straddle state boundaries!

----

#### _Compute Estimates of Forest Variables_

Now that you have loaded your FIA data into R, it's time to put it to work. Let's explore the basic functionality of `rFIA` with `tpa`, a function to compute tree abundance estimates (TPA, BAA, & relative abundance) from FIA data, and `fiaRI`, a subset of the FIA Database for Rhode Island including all inventories up to 2017.  

**Estimate the abundance of live trees in Rhode Island:**
```{r warning= FALSE, message=FALSE}
library(rFIA)
## Load the Rhode Island subset of the FIADB (included w/ rFIA)
## NOTE: This object was produced using readFIA
data("fiaRI")

## All Inventory Years Available (i.e., returns a time series)
tpaRI <- tpa(fiaRI)
## 2005 marks the first reporting year of the annual inventory, prior data is sparse
head(tpaRI[tpaRI$YEAR >= 2005,]) 

## Only estimates for the most recent inventory year
fiaRI_MR <- clipFIA(fiaRI, mostRecent = TRUE) ## subset the most recent data
tpaRI_MR <- tpa(fiaRI_MR)
head(tpaRI_MR)
```

**What if I want to group estimates by species? How about by size class?**
```{r warning= FALSE, message=FALSE}
## Group estimates by species
tpaRI_species <- tpa(fiaRI_MR, bySpecies = TRUE)
head(tpaRI_species)

## Group estimates by size class
## NOTE: Default 2-inch size classes, but you can make your own using makeClasses()
tpaRI_sizeClass <- tpa(fiaRI_MR, bySizeClass = TRUE)
head(tpaRI_sizeClass)

```

**What if I want estimates for a specific type of tree (ex. greater than 12-inches DBH and in a canopy dominant or subdominant position) in specific area (ex. growing on mesic sites), and I want to group by estimates by some variable other than species or size class (ex. ownsership group)?** Easy! Each of these specifications are described in the FIA Database, and all `rFIA` functions can leverage these data to easily implement complex queries!

``` {r warning = FALSE, message = FALSE}
## grpBy specifies what to group estimates by (just like species and size class above)
## treeDomain describes the trees of interest, in terms of FIA variables 
## areaDomain, just like above,describes the land area of interest
tpaRI_own <- tpa(fiaRI_MR, 
                     grpBy = 'OWNGRPCD', 
                     treeDomain = DIA > 12 & CCLCD %in% c(1,2),
                     areaDomain = PHYSCLCD %in% 20:29)
head(tpaRI_own)
```


**What if I want to produce estimates within my own population boundaries (within user-defined spatial zones/polygons)?** This is where things get really exciting. 
``` {r warning = FALSE, message = FALSE}
## Load the county boundaries for Rhode Island
data('countiesRI') ## Load your own spatial data from shapefiles using readOGR() (rgdal)

## polys specifies the polygons (zones) where you are interested in producing estimates
## returnSpatial = TRUE indicates that the resulting estimates will be joined with the 
##    polygons we specified, thus allowing us to visualize the estimates across space
tpaRI_counties <- tpa(fiaRI_MR, polys = countiesRI, returnSpatial = TRUE)

## NOTE: Any empty polygons below simply means no FIA data was available for that region
plot(tpaRI_counties['BAA']) # Plotting method for sf objects, also try 'TPA' or 'TPA_PERC'
```

**We produced a really cool time series earlier, how would I marry the spatial and temporal capacity of `rFIA` to produce estimates across user-defined polygons and through time?** Easy! Just hand `tpa` the full FIA.Database object you produced with `readFIA` (not the most recent subset produced with `clipFIA`). For stunning space-time visualizations, hand the output of `tpa` to `animateFIA`. To save the animation as a .gif file, simpy specify `fileName` (name of output file) and `savePath` (directory to save file, combined with `fileName`). 
```{r warning = FALSE, message=FALSE}
## Using the full FIA dataset, all available inventories
tpaRI_st <- tpa(fiaRI, polys = countiesRI, returnSpatial = TRUE)

## Animate the output
animateFIA(tpaRI_st, 'TPA', title = 'Abundance (TPA)', legend.height = .8)


```











